---
title: "Use `mrgsolve` to replace NONMEM simulations"
author: "Kenta Yoshida"
date: "`r Sys.Date()`"
output:  
  html_document:
    theme: cosmo
    code_folding: hide
    keep_md: yes
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

# Setup

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(tidyverse)
library(glue)
library(xpose)
library(mrgsolve)
library(xgxr)
library(tidyvpc)

theme_set(theme_linedraw(base_size = 12))

geomean <- function(x){exp(mean(log(x)))}
quibble <- function(x, q = c(0.05, 0.5, 0.95)) {
  tibble("{{ x }}" := quantile(x, q), 
         "{{ x }}_q" := paste0("P",
                               formatC(q * 100, width = 2, format = "d", flag = "0")))
}
# summarize(data, quibble(x))
```


# Preface

See <https://github.com/Genentech/mrgsolve_replace_NMsim> for the code & files needed to run the examples in your environment

## `mrgsolve` applications covered in this document

In this file, the following four types of `mrgsolve` applications has been demonstrated:

- Draw smooth PK profiles by adding observation time points in simulations
- Clinical trial simulation with new dose amount and schedules
- Obtain summary PK metrics (e.g. AUC<sub>ss</sub>, C<sub>min,ss</sub>, C<sub>min,ss</sub>) based on protocol-defined dosing schedules and empirical Bayes estimates (EBE) of individual ETA
  - This is useful for generating metrics for E-R analysis
- VPC


## Notes

- The model & data were from:
  - *Nonlinear Mixed‐Effects Model Development and Simulation Using nlmixr and Related R Open‐Source Packages*, Fidler et al., CPT Pharmacometrics Syst Pharmacol. 2019 Sep;8(9):621-633.
  - https://doi.org/10.1002/psp4.12445
- `mrgsolve` version >= 0.10.4 is required
- In these workflows, THETA (& SIGMA) is updated with values in input data frames, while OMEGA is updated with omat function to be applied on `mrgmod` object

# Preparations

## CSV data import
### First modify data

examplomycin.csv is obtained from Supplementary File 1 of the above article  
Modification needed to be compatible with NONMEM  

```{r, eval=FALSE}
df.raw <- read_csv("data/examplomycin.csv")

unique(df.raw$EVID)

df.raw %>% 
  mutate(CMT = ifelse(EVID == 101, 1, 2),
         EVID = ifelse(EVID == 101, 1, 0),
         PHASE = ifelse(TIME < 24, "SD", "MD")) %>% 
  relocate(CMT, .after = DV) %>% 
  rename(NMID = ID) %>% 
  write_csv("data/examplomycin2.csv")

```

### Import modified data

```{r}
df.nm <- read_csv("data/examplomycin2.csv")
df.nm.dose<- filter(df.nm, EVID == 1)
df.nm.obs <- filter(df.nm, EVID == 0)

time_units_dataset = "hours"
time_units_plot    = "days"
```

## Load model files

### `mrgsolve`

```{r}
mod_pk <- mread("mrgsolve/examplomycin_2cmt_covs_2020-07-08.cpp")
see(mod_pk)
```


### NONMEM fit

This is needed to extract population parameter estimates (THETA, OMEGA) as well as individuals (ETA).  

Alternatively you can manually modify population parameters in the `mrgsolve` model file.  
For ETA you most likely need to load NONMEM fit (unless you get EBE using R).  

```{r}
xpdb <- xpose_data(runno = "006", dir = "nm")

df.theta <- 
  get_prm(xpdb, transform = FALSE, quiet = TRUE) %>% 
  filter(type == "the") %>% 
  select(name, value) %>% 
  pivot_wider() %>% 
  rename(TVCL = THETA1,
         TVV2 = THETA2,
         TVV3 = THETA3,
         TVQ  = THETA4,
         TVKA = THETA5,
         PROP = THETA6,
         ADD  = THETA7,
         WTCL = THETA11,
         SEXV2= THETA12)

df.eta <- 
  get_data(xpdb, quiet = TRUE) %>% 
  select(NMID, 
         E_CLi = ETA1, 
         E_V2i = ETA2, 
         E_V3i = ETA3, 
         E_Qi  = ETA4, 
         E_KAi = ETA5) %>% 
  distinct()

omega.vec <- 
  get_prm(xpdb, transform = FALSE, quiet = TRUE) %>% 
  filter(type == "ome") %>% 
  pull(value)

d.omat <- 
  omat(group1 = bmat(omega.vec[1:3]), group2 = dmat(omega.vec[4:6]))
```



# Draw smooth individual PK curve

See https://mrgsolve.github.io/user_guide/topics.html#topic-designs for more details of specifying sampling designs

## Run

```{r}

# Set-up datasets
df.sim.act.dose <-
  df.nm %>% 
  # Add parameter values
  expand_grid(df.theta) %>% 
  mutate(ID = NMID)

df.sim.act.dose.eta <-
  df.nm %>% 
  # Add parameter and ETA values
  expand_grid(df.theta) %>% 
  left_join(df.eta, by = "NMID") %>% 
  mutate(ID = NMID)

# Model run setting
mod_pk.act.dose <- 
  mod_pk %>% 
  # Set Omega and Sigma to zero
  zero_re() %>% 
  # Keep these variables in the output
  # a.u.g is a flag for augmented sampling timepoints
  carry_out(NMID, AMT, EVID, WT, SEX, a.u.g) %>% 
  # Specify that we want to add augmented timepoints (defined with delta)
  obsaug()


# PRED 
## IPRED in output here is PRED because E_XXi is left at 0 and 
## SIGMA and OMEGA were also set as 0
mrgsim.pred.act.dose <- 
  mod_pk.act.dose %>% 
  data_set(df.sim.act.dose) %>%
  mrgsim(end = 200, delta = 1,
         recover = "PHASE")

df.mrgsim.pred.act.dose <- 
  as_tibble(mrgsim.pred.act.dose) %>% 
  rename(PRED = IPRED)

# IPRED
mrgsim.ipred.act.dose <- 
  mod_pk.act.dose %>% 
  data_set(df.sim.act.dose.eta) %>%
  mrgsim(end = 200, delta = 1,
         recover = "PHASE")

df.mrgsim.ipred.act.dose <- 
  as_tibble(mrgsim.ipred.act.dose)

```

## Plot

```{r, fig.width=6, fig.height=6}
df.mrgsim.ipred.act.dose %>% 
  ggplot(aes(TIME, IPRED)) +
  geom_line() +
  geom_line(data = df.mrgsim.pred.act.dose,
            aes(y = PRED),
            linetype = "dashed") +
  geom_point(data = df.nm.obs,
             aes(y = DV)) +
  # Only show the first 12 subjects
  facet_wrap_paginate(~NMID, nrow = 4, ncol = 4, page = 1,
                      labeller = "label_both") +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  labs(y = "Concentration",
       caption = "Solid line: IPRED, Dashed line: PRED, Circles: Obs")
```



# Simulate new dose or schedules

Regular NONMEM-format data frame (which you would export to CSV & use in NONMEM $SIM) would likely work as it is.  
If you want to utilize `mrgsolve`'s helper functions to generate datasets, see https://mrgsolve.github.io/user_guide/data-set-chapter.html#creating-data-sets for details.  

## Dose ranging
### Generate sim data

```{r}
set.seed(1234)

n.subj.per.dose <- 1000
doses <- c(600, 1200, 2400)
dosetime <- (0:14) * 24

n.subj.all <- n.subj.per.dose * length(doses)

# Generate dosing records

df.dose <- 
  tibble(AMT = rep(doses, each = n.subj.per.dose)) %>% 
  mutate(NMID = row_number()) %>% 
  mutate(CMT = 1, EVID = 1) %>% 
  expand_grid(TIME = dosetime)

# Here we are resampling covariates from the observed data
df.cov <- 
  df.nm %>% 
  select(NMID, WT, SEX) %>% 
  distinct() %>% 
  slice_sample(n = n.subj.all, replace = TRUE) %>% 
  mutate(NMID = row_number())


df.sim.dose.range <- 
  full_join(df.dose, df.cov, by = "NMID") %>% 
  # Add parameter values
  expand_grid(df.theta) %>%
  mutate(ID = NMID,
         NAMT = AMT)
```


### Run

```{r}
mrgsim.dose.range <- 
  mod_pk %>% 
  carry_out(NMID, NAMT, AMT, EVID, WT, SEX) %>% 
  data_set(df.sim.dose.range) %>%
  omat(d.omat) %>% 
  mrgsim(end = max(dosetime) + 24, delta = 1)

df.mrgsim.dose.range <- 
  as_tibble(mrgsim.dose.range) %>% 
  filter(EVID == 0) %>% 
  mutate(NAMT = factor(NAMT))
```


### Plot

```{r, fig.width=6, fig.height=3}
df.mrgsim.dose.range.qtile <- 
  df.mrgsim.dose.range %>% 
  group_by(NAMT, TIME) %>% 
  summarize(quibble(IPRED),
            n = n(),
            .groups = "drop") %>% 
  pivot_wider(names_from = IPRED_q, values_from = IPRED)

df.mrgsim.dose.range.qtile %>% 
  ggplot(aes(TIME, P50, color = NAMT)) +
  geom_line() +
  geom_ribbon(aes(ymax = P95, ymin = P05, fill = NAMT),
              alpha = .3, color = NA) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  facet_wrap(~NAMT, labeller = "label_both", nrow = 1) +
  labs(title = "Simulated PK with three dose levels",
       y = "Concentration")
```

### Generate exporure metrics

This can be used for showing PK variability or as input for E-R predictions

```{r}
exposure.metrics.dose.range <- 
  df.mrgsim.dose.range %>% 
  filter(TIME >= max(dosetime),
         TIME <= max(dosetime) + 24) %>% 
  group_by(NMID, NAMT) %>% 
  summarize(AUCss = max(AUCCENT) - min(AUCCENT),
            Cmaxss = max(IPRED),
            Cminss = min(IPRED),
            .groups = "drop")

exposure.metrics.dose.range

exposure.metrics.qtile <- 
  exposure.metrics.dose.range %>% 
  pivot_longer(AUCss:Cminss, names_to = "metric") %>% 
  group_by(NAMT, metric) %>% 
  summarize(quibble(value), .groups = "drop") %>% 
  pivot_wider(names_from = value_q, values_from = value)

exposure.metrics.qtile

exposure.metrics.dose.range %>% 
  ggplot(aes(AUCss, NAMT, fill = NAMT, color = NAMT)) +
  ggridges::geom_density_ridges(alpha = 0.5) +
  scale_y_discrete(expand = c(0, 0)) +
  ggridges::theme_ridges()
```



## Different treatment intervals

You can use ii and addl statements like NONMEM

### Generate sim data

```{r}
set.seed(1234)

n.subj.per.dose <- 1000
df.dose.interval.setting <- 
  tribble(~AMT, ~II, ~ADDL, ~NDOSE,
          600,  12,  29,    "600mg BID",
          1200, 24,  14,    "1200mg QD")
time.sim.end <- 15 * 24

n.subj.all <- n.subj.per.dose * nrow(df.dose.interval.setting)

# Generate dosing records

df.dose <- 
  df.dose.interval.setting %>% 
  expand_grid(IDdummy = 1:n.subj.per.dose) %>% 
  select(-IDdummy) %>% 
  mutate(NMID = row_number(),
         TIME = 0, CMT = 1, EVID = 1)

# Here we are resampling covariates from the observed data
df.cov <- 
  df.nm %>% 
  select(NMID, WT, SEX) %>% 
  distinct() %>% 
  slice_sample(n = n.subj.all, replace = TRUE) %>% 
  mutate(NMID = row_number())


df.sim.dose.intervals <- 
  full_join(df.dose, df.cov, by = "NMID") %>% 
  # Add parameter values
  expand_grid(df.theta) %>%
  mutate(ID = NMID)
```


### Run

```{r}
mrgsim.dose.intervals <- 
  mod_pk %>% 
  carry_out(NMID, AMT, EVID, WT, SEX) %>% 
  data_set(df.sim.dose.intervals) %>%
  omat(d.omat) %>% 
  mrgsim(end = time.sim.end, delta = 1,
         recover = "NDOSE")

df.mrgsim.dose.intervals <- 
  as_tibble(mrgsim.dose.intervals) %>% 
  filter(EVID == 0)
```


### Plot

```{r}
df.mrgsim.dose.intervals.qtile <- 
  df.mrgsim.dose.intervals %>% 
  group_by(NDOSE, TIME) %>% 
  summarize(quibble(IPRED),
            n = n(),
            .groups = "drop") %>% 
  pivot_wider(names_from = IPRED_q, values_from = IPRED)

df.mrgsim.dose.intervals.qtile %>% 
  ggplot(aes(TIME, P50, color = NDOSE)) +
  geom_line() +
  geom_ribbon(aes(ymax = P95, ymin = P05, fill = NDOSE),
              alpha = .3, color = NA) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  # facet_wrap(~DFREQ, labeller = "label_both", nrow = 1) +
  labs(title = "Simulated PK with two dosing intervals",
       y = "Concentration")
```



# Calc individual PK metrics

Sometimes we want to generate PK metrics for each individual (e.g. AUC<sub>ss</sub>, C<sub>min,ss</sub>, C<sub>min,ss</sub>) with empirical Bayes estimates (EBE), using protocol-defined dosing schedules rather than actual dosing records.  
This may be preferred in some situations where we want to avoid post-randomization bias.  


## Run

```{r}
# 15-day QD dosing of 1200mg
df.sim.prot.dose <- 
  df.nm %>% 
  select(NMID, WT, SEX) %>% 
  distinct() %>% 
  mutate(TIME = 0, AMT = 1200, II = 24, ADDL = 14, CMT = 1, EVID = 1) %>% 
  # Add parameter and ETA values
  expand_grid(df.theta) %>% 
  left_join(df.eta, by = "NMID") %>% 
  mutate(ID = NMID)

time.sim.end <- 16 * 24


# Model run setting
mod_pk.prot.dose <- 
  mod_pk %>% 
  # Set Omega and Sigma to zero
  zero_re() %>% 
  # Keep these variables in the output
  # a.u.g is a flag for augmented sampling timepoints
  carry_out(NMID, AMT, EVID, WT, SEX, a.u.g) %>% 
  # Specify that we want to add augmented timepoints (defined with delta)
  obsaug()


# IPRED
mrgsim.ipred.prot.dose <- 
  mod_pk.prot.dose %>% 
  data_set(df.sim.prot.dose) %>%
  mrgsim(end = time.sim.end, delta = 1)

df.mrgsim.ipred.prot.dose <- 
  as_tibble(mrgsim.ipred.prot.dose)

```


## Exporure metrics for individual subjects

This can be used for E-R analysis

```{r, fig.width=3, fig.height=5}
# You can use PKNCA::pk.calc.auc if the mrgsolve model does not calculate AUC in ODE
# exposure.metrics.ipred.prot.dose <- 
#   df.mrgsim.ipred.prot.dose %>% 
#   filter(TIME >= 15 * 24) %>% 
#   mutate(TAD = TIME -15 * 24) %>% 
#   group_by(NMID) %>% 
#   summarize(AUCss = PKNCA::pk.calc.auc(IPRED, TAD),
#             Cmaxss = max(IPRED),
#             Cminss = min(IPRED),
#             .groups = "drop")

# If mrgsolve calculate cumulative AUC, then max(AUC) - min(AUC)
# is AUC for the time interval
exposure.metrics.ipred.prot.dose <- 
  df.mrgsim.ipred.prot.dose %>% 
  filter(TIME >= 15 * 24,
         TIME <= 16 * 24) %>% 
  group_by(NMID) %>% 
  summarize(AUCss = max(AUCCENT) - min(AUCCENT),
            Cmaxss = max(IPRED),
            Cminss = min(IPRED),
            .groups = "drop")

exposure.metrics.ipred.prot.dose.plot <- 
  exposure.metrics.ipred.prot.dose %>% 
  pivot_longer(Cmaxss:Cminss) %>% 
  mutate(x = rep(1:2, times = nrow(exposure.metrics.ipred.prot.dose)),
         xj = jitter(x, amount = 0.09)) %>% 
  select(-AUCss, -x)

exposure.metrics.ipred.prot.dose.plot %>% 
  pivot_wider(names_from = name,
              values_from = c(value, xj)) %>% 
  ggplot() +
  geom_line(data = exposure.metrics.ipred.prot.dose.plot,
            aes(x = xj, y = value, group = NMID),
            alpha = 0.3) +
  geom_point(aes(xj_Cmaxss, value_Cmaxss), color = "dodgerblue", size = 2, alpha = 0.6) +
  geom_point(aes(xj_Cminss, value_Cminss), color = "darkorange", size = 2, alpha = 0.6) +
  expand_limits(y = 0) +
  scale_x_continuous(breaks = 1:2,
                     labels = c("Cmax", "Cmin"),
                     limits = c(0.5, 2.5)) +
  labs(x = "Individual exposure metrics",
       y = "")

exposure.metrics.ipred.prot.dose
```


# VPC

Here I used `tidyvpc` package

## Run sim

```{r}
# Sim for VPC
n.vpc <- 1000

df.sim.vpc <-
  df.nm %>% 
  # Add parameter values
  expand_grid(df.theta) %>% 
  # Duplicate all records n.vpc times
  # Need to give different ID#
  expand_grid(REP = 1:n.vpc) %>% 
  mutate(ID = NMID * n.vpc + REP) %>% 
  arrange(REP, NMID, TIME, EVID)

mrgsim.vpc <- 
  mod_pk %>% 
  carry_out(REP, NMID, AMT, EVID, WT, SEX) %>% 
  data_set(df.sim.vpc) %>%
  omat(d.omat) %>% 
  mrgsim(tad = TRUE, recover = "PHASE")

df.mrgsim.vpc <- 
  as_tibble(mrgsim.vpc) %>% 
  filter(EVID == 0)


# Sim PRED (required for tidyvpc)
df.sim.pred.for.vpc <-
  df.nm %>% 
  # Add parameter values
  expand_grid(df.theta) %>% 
  mutate(ID = NMID) 

mrgsim.pred.for.vpc <- 
  mod_pk %>% 
  carry_out(NMID, AMT, EVID) %>% 
  data_set(df.sim.pred.for.vpc) %>%
  zero_re() %>% 
  mrgsim(tad = TRUE, recover = "PHASE")

df.tad.pred <- 
  mrgsim.pred.for.vpc %>% 
  filter(EVID == 0) %>% 
  # Here IPRED is PRED as zero_re was used
  select(tad, PRED = IPRED)
```


## Generate VPC

```{r}
vpc <- 
  df.nm.obs %>% 
  bind_cols(df.tad.pred) %>% 
  mutate(PHASE = factor(PHASE, levels = c("SD", "MD"))) %>% 
  observed(x=TIME, y=DV) %>%
  simulated(df.mrgsim.vpc, y=DV) %>%
  stratify(~PHASE) %>% 
  # binning(bin = NTIM) %>% 
  binning(bin = "breaks", breaks = c(0, 1, 4, 5, 12, 24, 171, 185, 200)) %>% 
  vpcstats()

plot(vpc) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot,
                         breaks = (0:40) * 6)
```


## VPC with TAD

We can use time after dose (TAD) calculated in `mrgsim()` and use that for VPC plot, even if you don't have TAD in your original dataset

```{r}
vpc <- 
  df.nm.obs %>% 
  bind_cols(df.tad.pred) %>% 
  mutate(PHASE = factor(PHASE, levels = c("SD", "MD"))) %>% 
  observed(x=tad, y=DV) %>%
  simulated(df.mrgsim.vpc, y=DV) %>%
  binning(bin = "breaks", breaks = c(0, 1, 4, 5, 12, 22, 30)) %>%
  predcorrect(pred=PRED) %>%
  vpcstats()

plot(vpc) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot,
                         breaks = (0:40) * 6) +
  labs(x = "Time after the last dose (Days)")
```


# Session info

```{r}
devtools::session_info()
```


