---
title: "Make sure mrgsolve model replicate NONMEM sim"
author: "Kenta Yoshida"
date: "`r Sys.Date()`"
output:  
  html_document:
    theme: cosmo
    code_folding: hide
    keep_md: yes
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

# Setup

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(tidyverse)
library(glue)
library(xpose)
library(mrgsolve)
library(xgxr)
library(tidyvpc)

theme_set(theme_linedraw(base_size = 12))

geomean <- function(x){exp(mean(log(x)))}
quibble <- function(x, q = c(0.05, 0.5, 0.95)) {
  tibble("{{ x }}" := quantile(x, q), 
         "{{ x }}_q" := paste0("P",
                               formatC(q * 100, width = 2, format = "d", flag = "0")))
}
# summarize(data, quibble(x))
```


# Notes

Validation worked fine

- The model & data were from:
  - *Nonlinear Mixed‐Effects Model Development and Simulation Using nlmixr and Related R Open‐Source Packages*, Fidler et al., CPT Pharmacometrics Syst Pharmacol. 2019 Sep;8(9):621-633.
  - https://doi.org/10.1002/psp4.12445
- `mrgsolve` version >= 0.10.4 is required

# Import data

## First modify data

```{r, eval=FALSE}
df.raw <- read_csv("data/examplomycin.csv")

unique(df.raw$EVID)

df.raw %>% 
  mutate(CMT = ifelse(EVID == 101, 1, 2),
         EVID = ifelse(EVID == 101, 1, 0),
         PHASE = ifelse(TIME < 24, "SD", "MD")) %>% 
  relocate(CMT, .after = DV) %>% 
  rename(NMID = ID) %>% 
  write_csv("data/examplomycin2.csv")

```

## Import modified data

```{r}
df.nm <- read_csv("data/examplomycin2.csv")
df.nm.dose<- filter(df.nm, EVID == 1)
df.nm.obs <- filter(df.nm, EVID == 0)

time_units_dataset = "hours"
time_units_plot    = "days"
```

# Load model file

```{r}
mod_pk <- mread("mrgsolve/examplomycin_2cmt_covs_2020-07-08.cpp")
```

# Import NONMEM fit

```{r}
xpdb <- xpose_data(runno = "006", dir = "nm")

df.theta <- 
  get_prm(xpdb, transform = FALSE, quiet = TRUE) %>% 
  filter(type == "the") %>% 
  select(name, value) %>% 
  pivot_wider() %>% 
  rename(TVCL = THETA1,
         TVV2 = THETA2,
         TVV3 = THETA3,
         TVQ  = THETA4,
         TVKA = THETA5,
         PROP = THETA6,
         ADD  = THETA7,
         WTCL = THETA11,
         SEXV2= THETA12)

df.eta <- 
  get_data(xpdb, quiet = TRUE) %>% 
  select(NMID, 
         E_CLi = ETA1, 
         E_V2i = ETA2, 
         E_V3i = ETA3, 
         E_Qi  = ETA4, 
         E_KAi = ETA5) %>% 
  distinct()

omega.vec <- 
  get_prm(xpdb, transform = FALSE, quiet = TRUE) %>% 
  filter(type == "ome") %>% 
  pull(value)

d.omat <- 
  omat(group1 = bmat(omega.vec[1:3]), group2 = dmat(omega.vec[4:6]))
```

# Sim with actual dosing record

## Run

```{r}
df.sim.actdose <-
  df.nm %>% 
  # Add parameter and ETA values
  expand_grid(df.theta) %>% 
  left_join(df.eta, by = "NMID") %>% 
  rename(ID = NMID)


mod_pk.actdose <- 
  mod_pk %>% 
  zero_re() %>% 
  carry_out(AMT, EVID, WT, SEX, a.u.g) %>% 
  obsaug()

mrgsim.pred.actdose <- 
  mod_pk.actdose %>% 
  data_set(select(df.sim.actdose, !starts_with("E_"))) %>%
  mrgsim(end = 200, delta = 1,
         recover = "PHASE")

mrgsim.ipred.actdose <- 
  mod_pk.actdose %>% 
  data_set(df.sim.actdose) %>%
  mrgsim(end = 200, delta = 1,
         recover = "PHASE")


df.mrgsim.pred.actdose <- 
  as_tibble(mrgsim.pred.actdose) %>% 
  rename(NMID = ID,
         PRED = IPRED)

df.mrgsim.ipred.actdose <- 
  as_tibble(mrgsim.ipred.actdose) %>% 
  rename(NMID = ID)

```

## Plot

```{r}
df.mrgsim.ipred.actdose %>% 
  ggplot(aes(TIME, IPRED)) +
  geom_line() +
  geom_line(data = df.mrgsim.pred.actdose,
            aes(y = PRED),
            linetype = "dashed") +
  geom_point(data = df.nm.obs,
             aes(y = DV)) +
  # Only show the first 12 subjects
  facet_wrap_paginate(~NMID, nrow = 3, ncol = 4, page = 1,
                      labeller = "label_both") +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  labs(caption = "Solid line: IPRED, Dashed line: PRED, Circles: Obs")
```


## Compare between NONMEM and mrgsolve

```{r}
df.joined <- 
  bind_cols(df.mrgsim.ipred.actdose %>% 
              arrange(NMID, TIME) %>% 
              filter(a.u.g == 0, EVID == 0) %>% 
              select(NMID, TIME, IPRED_MRG = IPRED),
            get_data(xpdb, quiet = TRUE) %>%
              arrange(NMID, TIME) %>%  
              filter(EVID == 0) %>% 
              select(IPRED_NM = IPRED))

df.joined

qplot(IPRED_MRG, IPRED_NM, data = df.joined) +
  geom_abline(slope = 1, intercept = 0)
```

# DV comparison including VPC

## Run

### NONMEM

`vpc -samples=500 -auto_bin=auto -dir=vpc_run006_500 -predcorr -nodes=8 run006.mod`

```{r, eval=FALSE}

df.nm.vpc.output.raw <-
  read_table("nm/vpc_run006_500/m1/vpc_simulation.1.npctab.dta",
             col_types = cols(.default = col_double()),
             skip = 1) %>% 
  filter(!is.na(ID))

n.reps.vpc <- nrow(df.nm.vpc.output.raw) / nrow(df.nm)

df.nm.vpc.output <- 
  df.nm.vpc.output.raw %>% 
  mutate(REP = rep(1:n.reps.vpc, each = nrow(df.nm))) %>% 
  filter(MDV == 0)

save(df.nm.vpc.output, file = "nm/vpc/vpc.run006.Rdata")
```

```{r}
load("nm/vpc/vpc.run006.Rdata")

df.nm.vpc.output
```

### mrgsolve

```{r}
n.vpc <- 1000

df.sim.vpc <-
  df.nm %>% 
  # Add parameter values
  expand_grid(df.theta) %>% 
  rename(ID = NMID) %>% 
  # Duplicate all records n.vpc times
  expand_grid(REP = 1:n.vpc) %>% 
  mutate(NMID = ID,
         ID = ID * n.vpc + REP) %>% 
  arrange(REP, NMID, TIME, EVID)

mrgsim.vpc <- 
  mod_pk %>% 
  carry_out(REP, NMID, AMT, EVID, WT, SEX) %>% 
  data_set(df.sim.vpc) %>%
  omat(d.omat) %>% 
  mrgsim(recover = "PHASE")

df.mrgsim.vpc <- 
  as_tibble(mrgsim.vpc) %>% 
  filter(EVID == 0)
```

## Quantiles of DV for each individual

```{r}
df.nm.vpc.qtile <- 
  df.nm.vpc.output %>% 
  rename(NMID = ID) %>% 
  group_by(NMID, TIME) %>% 
  summarize(quibble(DV),
            n = n(),
            .groups = "drop")

df.vpc.qtile <- 
  df.mrgsim.vpc %>% 
  group_by(NMID, TIME) %>% 
  summarize(quibble(DV),
            n = n(),
            .groups = "drop")
```

```{r}
df.nm.vpc.qtile %>% 
  ggplot(aes(TIME, DV, group = DV_q, color = DV_q)) +
  geom_line() +
  geom_point(data = df.vpc.qtile) +
  coord_cartesian(xlim = c(0, 25)) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  facet_wrap_paginate(~NMID, nrow = 3, ncol = 4, page = 1,
                      labeller = "label_both") +
  labs(caption = "Lines: NONMEM, Points: mrgsolve")

df.nm.vpc.qtile %>% 
  ggplot(aes(TIME, DV, group = DV_q, color = DV_q)) +
  geom_line() +
  geom_point(data = df.vpc.qtile) +
  coord_cartesian(xlim = c(160, 200)) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  facet_wrap_paginate(~NMID, nrow = 3, ncol = 4, page = 1,
                      labeller = "label_both") +
  labs(caption = "Lines: NONMEM, Points: mrgsolve")

```

## Generate VPC

```{r}
d.pred <- 
  df.mrgsim.pred.actdose %>% 
  filter(a.u.g == 0, EVID == 0) %>% 
  pull(PRED)

vpc <- 
  df.nm.obs %>% 
  mutate(PRED = d.pred,
         NTIM = TIME,
         PHASE = factor(PHASE, levels = c("SD", "MD"))) %>% 
  observed(x=TIME, y=DV) %>%
  simulated(df.mrgsim.vpc, y=DV) %>%
  stratify(~PHASE) %>% 
  # binning(bin = NTIM) %>% 
  binning(bin = "breaks", breaks = c(1, 4, 5, 12, 24, 171, 185, 200)) %>% 
  vpcstats()

vpc.nm <- 
  df.nm.obs %>% 
  mutate(PRED = d.pred,
         NTIM = TIME,
         PHASE = factor(PHASE, levels = c("SD", "MD"))) %>% 
  observed(x=TIME, y=DV) %>%
  simulated(df.nm.vpc.output, y=DV) %>%
  stratify(~PHASE) %>% 
  # binning(bin = NTIM) %>% 
  binning(bin = "breaks", breaks = c(1, 4, 5, 12, 24, 171, 185, 200)) %>% 
  vpcstats()


plot(vpc) + coord_cartesian(ylim = c(0, 1000))
plot(vpc.nm) + coord_cartesian(ylim = c(0, 1000))
```

# IPRED comparison (with ETA resampling)

## Run

### NONMEM

```{r, eval=FALSE}

df.nmsim.output.raw <-
  read_table("nm/run006_sim.tab",
             col_types = cols(.default = col_double()),
             skip = 1) %>% 
  filter(!is.na(ID))

n.reps.vpc <- nrow(df.nmsim.output.raw) / nrow(df.nm)

df.nmsim.output <- 
  df.nmsim.output.raw %>% 
  mutate(REP = rep(1:n.reps.vpc, each = nrow(df.nm))) %>% 
  filter(EVID == 0)

save(df.nmsim.output, file = "nm/vpc/nmsim.run006_sim.Rdata")
```

```{r}
load("nm/vpc/nmsim.run006_sim.Rdata")

df.nmsim.output
```


## Quantile of IPRED for each individual

```{r}
df.nmsim.ipred.qtile <- 
  df.nmsim.output %>% 
  group_by(NMID, TIME) %>% 
  summarize(quibble(IPRED),
            n = n(),
            .groups = "drop")

df.ipred.qtile <- 
  df.mrgsim.vpc %>% 
  group_by(NMID, TIME) %>% 
  summarize(quibble(IPRED),
            n = n(),
            .groups = "drop")
```

```{r}
df.nmsim.ipred.qtile %>% 
  ggplot(aes(TIME, IPRED, group = IPRED_q, color = IPRED_q)) +
  geom_line() +
  geom_point(data = df.ipred.qtile) +
  coord_cartesian(xlim = c(0, 25)) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  facet_wrap_paginate(~NMID, nrow = 3, ncol = 4, page = 1,
                      labeller = "label_both") +
  labs(caption = "Lines: NONMEM, Points: mrgsolve")

df.nmsim.ipred.qtile %>% 
  ggplot(aes(TIME, IPRED, group = IPRED_q, color = IPRED_q)) +
  geom_line() +
  geom_point(data = df.ipred.qtile) +
  coord_cartesian(xlim = c(160, 200)) +
  xgx_scale_x_time_units(units_dataset = time_units_dataset, 
                         units_plot    = time_units_plot) +
  facet_wrap_paginate(~NMID, nrow = 3, ncol = 4, page = 1,
                      labeller = "label_both") +
  labs(caption = "Lines: NONMEM, Points: mrgsolve")

```



# Session info

```{r}
devtools::session_info()
```


